Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2012-03-19 pkg/paper/Paper/EmpEc.Rnw
small touch-ups
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@127 b679b2f7-6539-45d0-a605-5ddcb976cde9

Section~\ref{sec:fcrisis} focuses on the question of changes
in the inflation dynamics during the 2008/2009 financial crisis. Finally, the model fit and related issues will be discussed in Section~\ref{sec:fit}

@@

The first group that draws particular attention consists of the former
DEM (Deutsche Mark) zone countries, with the exception of Austria and
Luxembourg. All these countries had and still have a very low inflation rate,
both in terms of mean and variance, and their inflation dynamics
did not change significantly. This is plausible from an economic point of view
since the European central bank (ECB) is modeled after the Deutsche Bundesbank and the ECB did not
-- until recently -- diverge much from the monetary policy of the former. Thus,
the EMU did have no effect on inflation dynamics since the policies in
these countries were already much in line with Germany and hence the ECB. Table~\ref{tab:diag} does not 
suggest any problems in this segment.

@@

Austria, Finland, and France adjusted to the requirements of the EMU during the
first phase of convergence. However, with HAC correction, the very short segment of France would not be
segregated, thereby suggesting that France belongs to the group of countries which did not change at all.
In these countries,
the result of the adjustment process was a lower inflation rate and an overall
decrease in the variance.

The next group consists of three of the PIIGS countries and Luxembourg. Greece and
Portugal adjusted to the ERM shortly before or after its introduction.
Italy was successful in reducing its inflation rate, whereas Luxembourg
experienced a strong increase in its formerly very low inflation rate, both
in terms of mean and variance, also associated with a shift in skewness towards the
left. Apart from Luxembourg, the countries that had a structural break during the second 
phase of convergence were also successful in reducing mean inflation rates.
After the changes in this group, all countries have rather similar mean inflation
rates, which are closer to the levels of the previous two groups. The diagnostics for model fits in 
this group are unproblematic.

The fourth subgroup consists of two of the PIIGS countries, namely Ireland and Spain.  
There is a break in both countries in the summer of 2008 that captures the effects of
the financial crisis and the severe problems concerning the real estate situation.
In both countries, this had a strong deflationary impact, with deflation in Ireland
over an extended period of time. For the middle segment of Spain, the Ljung-box shows that not all
of the autocorrelation has been captured. In Spain, which was affected by a housing bubble of  
roughly the same size as Ireland, the housing prices did not adjust so abruptly and  
so strongly.

The next group consists of two countries that, after a very short participation in the
ERM, decided not to be part of it and hence also not to participate in the EMU:
the United Kingdom and Sweden. Both show clear breaks in the early 1990s which can be
traced back to economic crises: the currency crisis of the United Kingdom cumulating
into the ``Black Wednesday'' in 1992 and the banking crisis in Sweden during the early
1990s. With regard to the last segment for both countries, the Ljung-box suggests that the data 
in this segment is independently distributed. Anyhow, the HAC correction would also suggest two breaks 
for both series.

The last group consists of the Eastern European countries: Czech Republic,
Estonia, Hungary, Poland, Slovakia, and Slovenia. In almost all of these
countries, the mean and the variance of their inflation rates declined in the
late 1990s or early 2000s. For Estonia and Slovakia, the $\chi^2$ goodness-of-fit suggests that the generalized logistic distribution
might not be the most suitable distribution for this purpose. Also, compared to other country groups, the 
Eastern European group shows the highest occurrence of non-captured autocorrelation. Nevertheless, compared to other 
inflation rates, these series show the most dynamic behaviour.

Czech Republic, Estonia, and Hungary experienced a break in 1997/1998
resulting from efforts to curb in inflation by decreasing the growth of the money 
supply. The timing of the break thus indicates the date when the the biggest
transitionally shocks on the way towards free market economies have been overcome. In Poland, this
transition took two more years. Slovakia and Slovenia are somewhat different: On the one hand,
they started with lower mean inflation rates and on the other hand, their inflation dynamics
both changed roughly one year prior to their entry into the ERM~II.

>>>>

Finally, Section~\ref{sec:fcrisis} focuses on the question of changes
in the inflation dynamics during the 2008/2009 financial crisis.

@@

The first group that draws particular attention consists of the former
DEM (Deutsche Mark) zone countries, with the exception of Austria and
Luxembourg. All these countries had and still have a very low inflation rate,
both in terms of mean and variance, and their inflation dynamics
did not change significantly. This is plausible from an economic point of view
since the European central bank (ECB) is modeled after the Deutsche Bundesbank and the ECB did not
-- until recently -- diverge much from the monetary policy of the former. Thus,
the EMU did have no effect on inflation dynamics since the policies in
these countries were already much in line with Germany and hence the ECB. Table~\ref{tab:diag} does not 
suggest any problems in this group of countries.

@@

Austria, Finland, and France adjusted to the requirements of the EMU during the
first phase of convergence. In these countries, the result of the adjustment process
was a lower inflation rate and an overall decrease in the variance.
Additionally, if a HAC correction is used, the change for France would be nonsignificant,
suggesting that France may also belong to the group of countries which did not change at all.

The next group consists of three of the PIIGS countries and Luxembourg. Greece and
Portugal adjusted to the ERM shortly before or after its introduction.
Italy was successful in reducing its inflation rate, whereas Luxembourg
experienced a strong increase in its formerly very low inflation rate, both
in terms of mean and variance, also associated with a shift in skewness towards the
left. Apart from Luxembourg, the countries that had a structural break during the second 
phase of convergence were also successful in reducing mean inflation rates.
After the changes in this group, all countries have rather similar mean inflation
rates, which are closer to the levels of the previous two groups. The diagnostics 
from Table~\ref{tab:diag} indicate good model fits with low autocorrelations.

The fourth subgroup consists of two of the PIIGS countries, namely Ireland and Spain.  
There is a break in both countries in the summer of 2008 that captures the effects of
the financial crisis and the severe problems concerning the real estate situation.
In both countries, this had a strong deflationary impact, with deflation in Ireland
over an extended period of time. In Spain, which was affected by a housing bubble of  
roughly the same size as Ireland, the housing prices did not adjust so abruptly and  
so strongly.  The model fit diagnostics appear to be reasonable except for the middle
segment of Spain, where the Ljung-box shows that not all of the autocorrelation has been captured.

The next group consists of two countries that, after a very short participation in the
ERM, decided not to be part of it and hence also not to participate in the EMU:
the United Kingdom and Sweden. Both show clear breaks in the early 1990s which can be
traced back to economic crises: the currency crisis of the United Kingdom cumulating
into the ``Black Wednesday'' in 1992 and the banking crisis in Sweden during the early
1990s. With regard to the last segment for both countries, the Ljung-box suggests that 
the autocorrelation, albeit low around 0.16, is significant. However, both the standard
and the HAC-corrected sup$\mathit{LM}$ lead to highly significant evidence for structural
breaks.

The last group consists of the Eastern European countries: Czech Republic,
Estonia, Hungary, Poland, Slovakia, and Slovenia. In almost all of these
countries, the mean and the variance of their inflation rates declined in the
late 1990s or early 2000s. 
Czech Republic, Estonia, and Hungary experienced a break in 1997/1998
resulting from efforts to curb in inflation by decreasing the growth of the money 
supply. The timing of the break thus indicates the date when the the biggest
transitionally shocks on the way towards free market economies have been overcome. In Poland, this
transition took two more years. Slovakia and Slovenia are somewhat different: On the one hand,
they started with lower mean inflation rates and on the other hand, their inflation dynamics
both changed roughly one year prior to their entry into the ERM~II. Thus,
compared to other inflation rates, these series show the most dynamic behaviour. However,
the models also do not fit as well as for the previous groups of countries.
There is the highest occurrence of non-captured autocorrelation and for Estonia
and Slovakia the $\chi^2$ goodness-of-fit test suggests that the generalized logistic distribution
might not fit sufficiently well in the first segments.
